---
title: "Assignment 3"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: assignment3.bib
---

Clear R environment
```{r}
rm(list = ls())
```

# Problem 1. Random number generation and Monte Carlo integration.

## a) Generate $n$ independent samples from distribution

In this problem the task is to sample $n$ independent samples from the
probability distribution 

$$
p(x) = \frac{2^{\frac{1}{4}}\Gamma \left( \frac{3}{4}\right)}{\pi} \exp
  \left( -\frac{x^4}{2} \right) , \qquad -\infty < x < \infty 
$$

One might wish to implement a regular Metropolis-Hastings random walk for
sampling this distribution, but there is one major disadvantage for using this 
approach: [@wiki:metropolisco]

The samples are correlated, even though over the long term they do correctly 
follow $p(x)$. This is due to the fact that given that you know the current
time step, you have a pretty good estimate of what the value of the next one is.
In short, the adjacent samples are autocorrelated.

Because of this and the task specifically asking for independent samples.
A regular random walk metropolis-hastings approach is not feasible, we then turn
to the independent metropolis-hastings approach:

In R:

```{r}
# general 1d independence metropolis hastings
oneD.IRWMH <- function(prob,
                       sigma=1.0,
                       theta1=0.0,
                       Nsamp=10000){
  res <- numeric(Nsamp)
  # allocate memory
  res[1] <- theta1
  # old importance weight
  wt.old <- prob(res[1])/dnorm(res[1], sd=sigma)
  Nacc <- 0
    for(i in 2:Nsamp){
      # proposal (note, independent of past)
      thetaStar <- rnorm(1, sd=sigma)
      # new importance weight
      wt.star <- prob(thetaStar)/dnorm(thetaStar, sd=sigma)
      # accept probability
      alpha <- min(1.0, wt.star/wt.old)
      # accept/reject
      if(runif(1) < alpha){
        res[i] <- thetaStar
        wt.old <- wt.star
        Nacc <- Nacc+1
      } else {
        res[i] <- res[i-1]
      }
    }
  return(res)
}

# pdf from mandatory pdf
pdf <- function(x) {
  return ((2^0.25*gamma(3/4)/pi) * exp(-x^4 / 2))
}


# sample n samples
n <- 10000
out <- oneD.IRWMH(pdf, theta1 = 0.0, sigma=1, Nsamp = n)

# plot 
par(mfrow=c(2,1))
plot(1:length(out),out,pch=20,cex=0.1,xlab="MCMC iteration #")
hist(out, probability = TRUE)
curve(pdf, add = TRUE, -2, 2)
```
## b) Generate $n$ independent samples from distribution

Now let's sample from the distribution

$$
p(x) = 2x \exp(-x^2), \qquad 0 < x < \infty
$$
This function seems to be quite simple to find the CDF for and direct sampling 
via inverse transform sampling is possible. We get

$$
F(x) = \int_{0}^{x} p(x) = 1 - e^{-x^2}
$$
And by the inverse sampling method, we can calculate 

$$
X = F^{-1}(U) = \pm \sqrt{-\ln(1 - u)}
$$
Given the nature of calculating square roots, we get positive and negative 
X values but since the support of the pdf clearly states that $0 < x < \infty$
we reject the negative values.

in R:

```{R}
# Use inverse sampling method to sample from
# p(x) = 2x exp(-x^2)
inverse_sampling <- function(Nsamples) {
  # Sample n samples from uniform distribution
  u <- runif(Nsamples)
  # Return X = F^-1(U)
  return(sqrt(-log(1-u)))
}

# PDF
pdf2 <- function(x) {
  return(2*x * exp(-x^2))
}

x <- inverse_sampling(10000)
hist(x, probability = TRUE)
curve(pdf2, 0, 3, add = TRUE)
```

## c) Evaluate integral using monte-carlo integration

Now we will consider the integral

$$
\int_0^\infty \exp(\sqrt{x})\exp(-20(x-4)^2) \, dx
$$
Let's evaluate the integral using importance sampling. First let's plot the 
function $g(x)$ that is being integrated:

In R:

```{r}
g <- function(x) {
 exp(sqrt(x))*exp(-20*(x-4)^2)
}

f <- function(x) {
  dnorm(x, mean=4, sd=0.1)
}

x<-seq(1,10,0.01); y1=g(x); y2=f(x)
plot(x, y1, type="l", pch=19, col="red", xlab="x", ylab="y")
lines(x, y2, pch=18, col="blue", type="l", lty=2)
legend(1, 95, legend=c("Line 1", "Line 2"),
       col=c("red", "blue"), lty=1:2, cex=0.8)
```
With the function we want to integrate in $g(x)$ in red and the $f(x)$ in 
dashed blue. We want to calculate $E\left(\frac{g(x)I(x \in A)}{f(x)}\right)$
by sampling $X$ values from $f$ and calculating the empirical mean of 
$\frac{g(x)I(x \in A)}{f(x)}$ where $I$ is the indicator function for the 
support.

```{r}
importance <- function(Nsamp) {
  # Sample from f
  x <- rnorm(Nsamp, 4, 0.1)
  # calculate empirical mean
  return(mean((g(x)*(x>0))/(f(x))))
}

# Estimate of integral using importance sampling
importance(100000)
# Numerical integration
integrate(g, 0, Inf)$value
```

# Problem 2. Smile shaped target

## a) Sampling using 2 random walk

In this exercise, we shall sample from the distribution

$$
\log g(\boldsymbol \theta) = -\frac{\theta_1^2}{2} - 
\frac{(\theta_2 - \theta_1^2)^2}{2}, \qquad \infty < x < \infty
$$
To do this, an 2D random walk with multivariate normal proposals as been
implemented in R.

In the code below, a 2D random walk with proposal

$$
N(\boldsymbol \theta, \Sigma)
$$
Where 

$$
\Sigma = 
\begin{bmatrix}
2, 0 \\
0, 2
\end{bmatrix}
$$
and
$$
\boldsymbol \theta = 
\begin{pmatrix}
0 \\
0
\end{pmatrix}
$$
Yielded the greatest effective sample size.

In R:

```{r}
# Load the multivariate normal library 
library(mvtnorm)
library(coda)

# Target function from 
logg <- function(theta) {
  return(-theta[1]^2/2 - (theta[2] - theta[1]^2)^2 /2)
}

# 2D Random walk using log g
smile_shaped <- function(logg,
                         sigma=diag(2),
                         theta=c(0.0, 0.0),
                         Nsamp=100000){
  
  # Reserve 2xNsamp matrix with zeros
  res <- matrix(0, Nsamp, 2)
  
  # Set initial conditions
  res[1,] <- theta
  
  # Calculate old log-probability
  logold <- logg(theta)
  
  # accept counter
  Nacc <- 0
  
  # Iterate no of samples.
  for(i in 2:Nsamp){
    # Proposal step
    new <- res[i-1,] + rmvnorm(1, theta, sigma)
    # Log-p of proposal step
    lognew <- logg(new)
    # Evaluate step
    logfrac <- exp(min(0.0,lognew-logold))
    # Accept or reject new step
    if(runif(1)<logfrac && is.finite(logfrac)){
      # accept
      res[i,] <- new
      logold <- lognew
      Nacc <- Nacc+1
    } else {
      # reject
      res[i,] <- res[i-1,]
    }
  }
  print(paste("Accept prob: ", Nacc/Nsamp))
  return(res[5000:Nsamp,])
}

# Covariance matrix
sigma <- matrix(c(4.5, 0, 0, 4.5), nrow=2, ncol=2)
m <- smile_shaped(logg=logg, sigma=sigma, c(0.0, 0.0), 50000)
```

Let's plot the ``Smile shaped'' distribution:

```{r}
# Plot multivariate distribution
plot(m)
```

Let's also see if the marginal distribution of $\theta_1$ is standard normal:

```{r}
# Plot marginal of theta1 and compare with standard normal
hist(m[,1], probability=TRUE)
curve(dnorm(x, mean=0, sd=1), 
      col="red", lwd=2, add=TRUE, yaxt="n")
```

To check if we have explored the distribution well, let's calculate the effective
sample size of $theta_1$ and $theta_2$.

```{r}
ESS <- function(x) {
  coda::effectiveSize(x)
}

ESS(m[,1])
ESS(m[,2])
```
# Problem 3. IMH for simple logistic regression problem

```{r}
# Load the data set
df <- data.frame(read.table("logistic_regression_data.txt"))
x <- df$x
y <- df$y

# function returning a log-posterior kernel for theta
logistic.lp <- function(theta) {
  alpha <- theta[1]
  beta <- theta[2]
  # log-likelihood
  Eeta <- exp(alpha + beta*x)
  p <- Eeta/(1.0 + Eeta)
  log.like <- sum(dbinom(y, size=1, prob = p, log=TRUE))
  
  # priors
  log.prior <- dnorm(alpha, sd=10, log=TRUE) + dnorm(beta, sd=10, log=TRUE)
  
  # log-posterior kernel
  return(log.like+log.prior)
}
```

## a) Sample $\alpha$ and $\beta$ using metropolis-hasting and bayesian inference

In this task, we will use independent metropolis-hastings sampling for the
posterior distribution $p(\boldsymbol \theta | \boldsymbol y)$ using a 
$N(\hat{\theta}, \delta \hat{\Sigma})$ proposal distribution.

In R:

```{r}
# Independent sampler
IndepMH <- function(lprob, theta=c(0, 0), sigma=diag(2), Nsamp=1000, S=1){
  # Allocate space
  res <- matrix(0, Nsamp, 2)
  # Set initial values
  res[1,] <- theta
  # Old importance
  pold <- lprob(theta) - dmvnorm(theta, mean=theta ,sigma=sigma, log=TRUE)
  # No of accept
  Nacc <- 0
  for (i in 2:Nsamp){
    # Sample theta from multivariate normal with scaled sigma
    new <- rmvnorm(1, theta, S*sigma)
    # Importance of new theta
    pnew <- lprob(new) - dmvnorm(new, mean=theta, sigma=sigma, log=TRUE)
    frac <- exp(min(0.0, pnew-pold))
    if(runif(1)<frac && is.finite(frac)){
      # Accept
      res[i,] <- new
      pold <- pnew
      Nacc <- Nacc+1
    } else {
      # Reject
      res[i,] <- res[(i-1),]
    }
  }
  print(paste0("accept rate : ",Nacc/Nsamp))
  return(res[1000:Nsamp,])
}

sigma <- matrix(c(0.00653,-0.00058,-0.00058,0.01689),2,2)

m <- IndepMH(logistic.lp, c(-0.102, 1.993), sigma, 10000, 0.6)
plot(m)
```

Calculate ESS

```{r}
ESS(m[,1])
ESS(m[,2])
```

b/c) Plot quantiles of data and find $x*$ such that $P(m(x*) > 0.8) = 0.99$

To  solve this problem, we calculate $m(x*)$ the median and quantiles for
$x \in [-5, 5]$ where $\alpha$ and $\beta$ has been estimated based on the
dataset provided in the mandatory assignment.

We use the following algorithm to solve this problem:

1. Given data, estimate $\alpha$, $\beta$ using independent metropolis hastings
2. Select sequence of $x*$ we want to explore
3. Iterate through each value of $x*$
4. Calculate median and 01th, 05th and 95th quantiles and return these as a vector
5. Store vector as row in matrix
6. Repeat steps 4-5 for each value in $x*$

When this algorithm terminates, we have a 2D matrix of medians and quantiles for
all $x*$. We use this for plotting and finding the $x*$ where 
$P(m(x*) > 0.8) = 0.99$

In R:

```{r}
# Plotdist is function for calculating values for plotting
plotdist <- function (x) {
  # Calculate m(x*)
  mstar <- exp(m[,1] + m[,2]*x) / (1 + exp(m[,1] + m[,2]*x))
  # calulcate median given x*
  med <- median(mstar)
  # Caltulate quantiles given x*
  quant05 <- quantile(mstar, 0.05)
  quant95 <- quantile(mstar, 0.95)
  quant01 <- quantile(mstar, 0.01)
  # return vector of calculations
  return(c(med, quant05, quant95, quant01))
}

# sequence of x* values
x <- seq(-5, 5, 0.01)
# plot 
plotmat <- matrix(0, length(x), 4)
# set index
i <- 0
# Loop through all x* values in sequence
for (val in x) {
  # increment index for each value in xÆ
  i <- i+1
  plotmat[i,] <- plotdist(val)
}

# First value larger than 0.8 (for plotting horizontal)
horizontal <- plotmat[, 4][plotmat[, 4] > 0.8][1]

# First x-value of m-value with 99% probability of being larger than 0.8
vertical <- x[Position(function(x) x > 0.8, plotmat[, 4])]

# Plot all medians and quantiles conditional on x*
plot(x, plotmat[, 1], type="l", col="red", ylab="M*")
lines(x, plotmat[, 2], type="l", col="blue", lty=2)
lines(x, plotmat[, 3], type="l", col="orange", lty=3)
lines(x, plotmat[, 4], type="l", col="green", lty=4)
# Add vertical line for x values
abline(v = vertical, col="black")
# add horizontal line for m values
abline(h = horizontal, col="black")
legend("topleft", legend=c("Median", "05 quantile", "95 quantile", "01 quantile"),
       col=c("red", "blue", "orange", "green"), lty=1:4, cex=0.8)

# X* where P(M(X*) > 0.8) = 0.99 
print(vertical)
```

# Problem 4. Gibbs sampler for simple linear regression model

## a) Find conditional posteriors.

The posterior for $\tau$ is given as

$$
\pi(\tau | \alpha, \beta, \boldsymbol{y}) \sim
\text{gamma}\left(\frac{n}{2} + 1, \frac{1}{\frac{1}{2}\sum_{i=1}^n
(y_i -\alpha - \beta x_i)^2 + 1
}\right)
$$

To calculate the posterior of $\alpha$ we see by inspection that the log-density
kernel is possibly on the form 
$\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) = a + b\alpha + c\alpha^2$ which 
tells us the distribution is normal and that the constant $a$ is unimportant.

Let's calculate $\log \pi(\alpha | \beta, \tau, \boldsymbol y)$.

From the log-density kernel in the assignment description, we get

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c'_\alpha - \frac{\tau}{2}\sum_{i=1}^n (y_i - \alpha - \beta x_i)^2 - 
\frac{\alpha^2}{200}.
$$
Where $c'_\alpha$ is a unimportant constant containing all values not dependent 
on $\alpha$. By squaring the term in the sum and adding unimportant values to
$c'_\alpha$ we now get.

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c'_\alpha - \frac{\tau}{2}(n \alpha^2 + \alpha \sum_{i=1}^n(2\beta x_i - 2y_i)) - 
\frac{\alpha^2}{200}
$$
By multiplying the $\tau$ term and reshuffling we get

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c_\alpha - \tau\sum_{i=1}^n(\beta x_i - y_i)\alpha -\left(\frac{\tau n}{2} + 
\frac{1}{200}\right)\alpha^2
$$
And we have

$$
b_\alpha = -\tau\sum_{i=1}^n(\beta x_i - y_i)
$$
and

$$
c_\alpha = -\left(\frac{\tau n}{2} + 
\frac{1}{200}\right)
$$

and the posterior (from the lecture notes) is

$$
\pi(\alpha | \beta, \tau, \boldsymbol{y}) \sim
N\left( -\frac{b_\alpha}{2c_\alpha},  -\frac{1}{2c_\alpha} \right)
$$

And we do the same for $\pi(\beta | \alpha, \tau, \boldsymbol{y})$

From the log-density kernel in the assignment description, we get

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \frac{\tau}{2}\sum_{i=1}^n (y_i - \alpha - \beta x_i)^2 - 
\frac{\beta^2}{200}.
$$
Where $c'_\beta$ is a unimportant constant containing all values not dependent 
on $\beta$. By squaring the term in the sum and adding unimportant values to 
$c'_\beta$ we now get.

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \frac{\tau}{2}\sum_{i=1 }^n \left(2\alpha\beta x_i + \beta^2 x_i^2 
- 2\beta x_i y_i
\right) - \frac{\beta^2}{200}
$$
By multiplying the $\tau$ term and reshuffling we get

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \tau\left(\sum_{i=1}^n \alpha x_i - \sum_{i=1}^n x_i y_i \right)\beta 
- \left(\frac{\tau}{2} \sum_{i=1}^n x_i^2 + \frac{1}{200}\right)\beta^2
$$
And we have

$$
b_\beta = 
- \tau\left(\sum_{i=1}^n \alpha x_i - \sum_{i=1}^n x_i y_i \right)
$$
and

$$
c_\beta = 
- \left(\frac{\tau}{2}\sum_{i=1}^n x_i^2 + \frac{1}{200}\right)
$$

and the posterior (from the lecture notes) is

$$
\pi(\beta | \alpha, \tau, \boldsymbol{y}) \sim
N\left( -\frac{b_\beta}{2c_\beta},  -\frac{1}{2c_\beta} \right)
$$

## b) Gibbs sampler with three blocks

Load the data

```{r}
# Load regression data
df <- data.frame(read.table(file="linear_regression_data.txt"))
x <- df$x
y <- df$y
``` 

Let's implement the gibbs sampler

```{r}
gibbs <- function(x, y, theta=c(1.0, -1.2, 1.0), Nsamp=10000) {
  # Result matrix (State stored thrice each iteration)
  res <- matrix(0.0, 3*Nsamp+1, 3)
  # Set state 1 to initial values
  res[1, ] <- theta
  # Set gamma shape parameter
  gammashape <- length(x)/2 + 1
  # Set counter
  k <- 2
  # Iterate...
  for(i in 1:Nsamp){
    
      # Block 1. Alpha
      balpha <- -theta[3]*sum(theta[2]*x - y)
      calpha <- -theta[3]*length(y)*0.5 - 1/200
      theta[1] <- rnorm(1, -balpha/(2*calpha), -1/(2*calpha))
      # store state
      res[k, ] <- theta
      k <- k+1
      
      # Block 2. Beta
      bbeta <- -theta[3]*(sum(x*theta[1]) - sum(x*y))
      cbeta <- -theta[3]*0.5*sum(x^2) - 1/200
      theta[2] <- rnorm(1, -bbeta/(2*cbeta), -1/(2*cbeta))
      # store state
      res[k, ] <- theta
      k <- k+1
      
       # Block 3. Tau
      gammascale <- 1/(0.5*sum((y - theta[1] - theta[2]*x)^2))
      theta[3] <- rgamma(1, gammashape, scale = gammascale)
      # store state
      res[k, ] <- theta
      k <- k+1
  }
  # Return results minus burn in
  return(res[500:k-1,])
}

# Get samples to matrix
m <- gibbs(x, y)
# Calculate column means
colMeans(m)
# Compre with one simple linear regression
lm(y~x, data=df)
```
## c) Check if samples have converged

Our means seem pretty good, let's plot trace plots of alpha and beta.

```{r}
# Traceplot alpha vs beta
plot(m[,1], m[,2], type="l")
```
Also, let's calculate all the effective sample sizes for each variable

```{r}
ESS(m[,1])
ESS(m[,2])
ESS(m[,3])
```
We see that we have a pretty good sample size. Now let's investigate whether we
can improve this by reducing the no of blocks and making blocks less dependent.

## d) Improve gibbs sampler

Let's combine the blocks for $\alpha$ and $\beta$ from two normal distribution
to one block of a multinormal distribution. With a posterior distribution
$\pi(\alpha, \beta | \tau, \boldsymbol y) \sim N(\boldsymbol \mu, \boldsymbol\Sigma)$ 

where

$$
\boldsymbol\Sigma = -\boldsymbol c^{-1}
$$
and

$$
\boldsymbol \mu = -\boldsymbol c^{-1}\boldsymbol b
$$
and $\boldsymbol c$ and $\boldsymbol b$ is given in the assignment.

Let's alter the function for gibbs sampling to have two blocks instead.

```{r}
gibbs_multi <- function(x, y, theta=c(1.0, -1.2, 1.0), Nsamp=10000) {
  # Result matrix (State stored thrice each iteration)
  res <- matrix(0.0, 2*Nsamp+1, 3)
  # Set state 1 to initial values
  res[1, ] <- theta
  # Set gamma shape parameter
  gammashape <- length(x)/2 + 1
  # Set counter
  k <- 2
  
  # Calculations that can be done once for improvment of performance
  # Sum of x
  sumx <- sum(x)
  # Sum of x^2
  sumx2 <- sum(x^2)
  # Sum of y
  sumy <- sum(y)
  # Sum of xy
  sumyx <- sum(x*y)
  # No of datapoints
  n <- length(x)
  
  # Iterate...
  for(i in 1:Nsamp){
    
    # Block 1. [alpha, beta]
    # C from assignment description
    c <- matrix(c(-n*theta[3] - 0.01, -theta[3]*sumx,
                  -theta[3]*sumx, -theta[3]*sumx2 - 0.01), nrow=2, ncol=2)
    # b from assignment description
    b <- c(theta[3]*sumy, theta[3]*sumyx)
    
    theta[1:2] <- rmvnorm(1, -solve(c)%*%b, -solve(c))
    res[k, ] <- theta
    k <- k+1
    
     # Block 2. Tau
    gammascale <- 1/(0.5*sum((y - theta[1] - theta[2]*x)^2))
    theta[3] <- rgamma(1, gammashape, scale = gammascale)
    # store state
    res[k, ] <- theta
    k <- k+1
  }
  # Return results minus burn in
  return(res[500:k-1,])
}

m2 <- gibbs_multi(x, y)
colMeans(m2)
plot(m2[,1], m2[,2], type = "l")
```
Now, is this gibbs sampler better than our first implementation? Let's check the
effective sample size:

```{r}
ESS(m2[,1])
ESS(m2[,2])
ESS(m2[,3])
```
Which makes sense, we sampled $beta$ and $alpha$ with normal distributions
dependent on the values of one another, in this case we sample both from a
multinormal distribution according to the given covariance matrix 
where each pair of samples are independent from one
another.

Fewer blocks and striving for independence among blocks will, according to the
lectures, give better samples.

## e) Investigate whether $\alpha + \beta = 0$

To do a simple ``hypothesis test'' using quantiles and our samples. for every
sample, calculate $\alpha + \beta$ and store the values, calculate the 95% 
quantile confidence interval and plot the histogram with the interval.

This could be done more appropriately by using a more analytical approach for
our calculations but we resort to this simple case.

```{r}
alphapbeta <- m[,1] + m[,2]
hist(alphapbeta, breaks=25, 
     main = "3-block gibbs sampler: alpha + beta distribution")
abline(v = quantile(alphapbeta, 0.975), col="red", lwd="2")
abline(v = quantile(alphapbeta, 1-0.975), col="red", lwd="2")
```

```{r}
alphapbeta2 <- m2[,1] + m2[,2]
hist(alphapbeta2, breaks=25,
     main = "2-block gibbs sampler: alpha + beta distribution")
abline(v = quantile(alphapbeta2, 0.975), col="red", lwd="2")
abline(v = quantile(alphapbeta2, 1-0.975), col="red", lwd="2")
```

In the first distribution, we accept (with a significance of 5%), that
$\alpha + \beta \neq 0$. 

In the second distribution, we fail to reject the null-hypothesis that  
$\alpha + \beta = 0$. 

# Problem 5. Simple linear regression analysis, bootstrap approach

Now, let's address the same linear regression problem as problem 4 but instead
of using bayesian inference, we consider a classical approach using 
bootstrapping. 

In our case, our statistic will be the simple linear regression on bootstrapped
datasets such that we get $\alpha$ and $\beta$ pairs for each bootstrapped datasets
and we can investigate their distribution.

Let's implement this in R:

```{r}
# Import bootstrap library
library(boot)

# Our statistic (simple linear regression y ~ alpha + beta x)
slr <- function(data,i) {
  lm(data[i, 1]~data[i, 2],data = df)$coefficients
}   

# Run bootstrap on our data
boot.obj <- boot(data = df, statistic = slr, R = 5000)

# Traceplot bootstrap estimate to compare with our bayesian approach
plot(boot.obj$t[,1], boot.obj$t[,2], main="Bootstrap traceplot", type="l")
```

We see that out bootstrapped joint distribution is very similar to the bayesian 
joint distribution that we calculated in problem 4. 

Now, let's go through each of the following cases

1. $\sigma = 1/\sqrt{\tau}$ is strictly smaller than 1.0 
2. $\alpha$ is equal to 0
3. $\alpha + \beta$ is equal to 0

```{r}
# Set gamma shape parameter
gammashape <- length(x)/2 + 1

# Calculate sigma for every alpha-beta pair
sigma <- 1/sqrt(
  rgamma(5000, shape = gammashape,
  scale = 1/(1/2 * sum((y - mean(boot.obj$t[,1]) - 
                        mean(boot.obj$t[,2])*x)^2) + 1))
)

# Claim 1
hist(sigma, breaks=25, main="Distribution of sigma")
abline(v = 1, col="red", lwd="2")

# Claim 2
hist(boot.obj$t[,1], breaks=25, main="Distribution of alpha")
abline(v = quantile(boot.obj$t[,1], 0.975), col="red", lwd="2")
abline(v = quantile(boot.obj$t[,1], 1-0.975), col="red", lwd="2")

# Claim 3
hist(boot.obj$t[,1] + boot.obj$t[,2], breaks=25, 
     main="Distribution of alpha + beta")
abline(v = quantile(boot.obj$t[,1] + boot.obj$t[,2], 0.975), 
       col="red", lwd="2")
abline(v = quantile(boot.obj$t[,1] + boot.obj$t[,2], 1-0.975), 
       col="red", lwd="2")
```
Using our quantile confidence intervals, we reject claim 1 as
it is clear that $tau$ is greater than 1 on several occasions. claim 2 is
also rejected, our 95% confidence interval does not include $\alpha = 0$. Claim 
3 holds in our since the confidence interval for $\alpha + \beta$ contains 0. 

# Problem 6. Bayesian inference for the SEIR model.

Let's once again take a look at the SEIR model. We are tasked with estimating
the reproduction number $R$ and the social distancing factor $S$ based on the
dataset ${y_t}_{i=1}^T$ of counts of the number of hospitalized persons (out
of a population of 1 million) recorded  every day between 
\formatdate{31}{03}{2020} and \formatdate{18}{07}{2020}.

We model the data using independent Poisson distributions

$$
y_y \sim \text{Poisson}(\lambda(t, \boldsymbol \theta)), \qquad t = 1, \dots, N
$$
where $\lambda(t, \boldsymbol \theta)$ is taken to be the combined 
(SEIR-model implied) number of persons in the hospital states HH, HC and CC 
(i.e. 1000000(HH +HC +CC)) for the days t we have data for.

## a) Obtain MCMC samples targeting $p(\boldsymbol \theta | \boldsymbol y)$

Let's run a 2D random walk on the posterior and see if we are able to converge
on some value. We choose the mean vector that we found in assignment 2 problem 
4 c for tuning the theta variable, the code and covariance matrix is borrowed 
from the tip in the assignment text.


```{r}
source("seir_mcmc_funs.R")

twoD_RWMH_poisson <- function(lprob, # log-probability density kernel
                     Sigma=diag(2), # default proposal covariance = identity matrix
                     theta1=c(3.18,0.22), # default initial configuration
                     n=1000){

  # Cholesky factorization (used for effective sampling)
  L = t(chol(Sigma))
  
  output <- matrix(0.0,n,2)
  output[1,] <- theta1
  # store old lprob
  pastVal <- lprob(theta1)
  # accept counter
  Nacc <- 0
  # main iteration loop
  for(i in 2:n){
    # proposal
    thetaStar <- output[(i-1),] + L%*%rnorm(2)
    currVal <- lprob(thetaStar)
    # accept prob
    alpha <- exp(min(0.0,currVal-pastVal))
    # accept/reject
    if(runif(1)<alpha && is.finite(alpha)){
      # accept
      output[i,] <- thetaStar
      pastVal <- currVal
      Nacc <- Nacc+1
    } else {
      output[i,] <- output[(i-1),]      
    }
  } # main iteration loop
  
   print(paste0("RWMH done, accept rate : ",Nacc/(n-1)))
  return(round(output,5))
  
}

# Tune model with mean vector from assignment 2
m <- twoD_RWMH_poisson(seir.lp, theta1=c(3.18,0.22),
                       Sigma=matrix(c(0.0003,0,0,0.0000005),2,2),
                       n=1000)

plot(m[,1], m[,2], type="l")
```

From our traceplot, we see that the random walk takes us from the
mean of the posterior and it converges down on $3.0 < R < 3.05$ and 
$S \approx 0.17$. Let's start our random walk with a mean vector closer to
these values and run the code again.

```{r}
# Tune model with mean vector from assignment 2
m <- twoD_RWMH_poisson(seir.lp, theta1=c(3.02,0.167),
                       Sigma=matrix(c(0.0003,0,0,0.0000005),2,2),
                       n=1000)

plot(m[,1], m[,2], type="l")
ESS(m[,1])
ESS(m[,2])
```
Still, our sample size is relatively small, let's use our independent random walk
implementation instead.

```{r}
# Independent sampler
IndepMH <- function(lprob, theta=c(0, 0), sigma=diag(2), Nsamp=1000, S=1){
  # Allocate space
  res <- matrix(0, Nsamp, 2)
  # Set initial values
  res[1,] <- theta
  # Old importance
  pold <- lprob(theta) - dmvnorm(theta, mean=theta ,sigma=sigma, log=TRUE)
  # No of accept
  Nacc <- 0
  for (i in 2:Nsamp){
    # Sample theta from multivariate normal with scaled sigma
    new <- rmvnorm(1, theta, S*sigma)
    # Importance of new theta
    pnew <- lprob(new) - dmvnorm(new, mean=theta, sigma=sigma, log=TRUE)
    frac <- exp(min(0.0, pnew-pold))
    if(runif(1)<frac && is.finite(frac)){
      # Accept
      res[i,] <- new
      pold <- pnew
      Nacc <- Nacc+1
    } else {
      # Reject
      res[i,] <- res[(i-1),]
    }
  }
  print(paste0("accept rate : ",Nacc/Nsamp))
  return(res)
}


m <- IndepMH(seir.lp, theta=c(3.02, 0.167),
             sigma=matrix(c(0.0003,0,0,0.0000005),2,2),
             Nsamp=1000, S=0.8)

plot(m[,1], m[,2], type="l")
ESS(m[,1])
ESS(m[,2])
```

Finally we have a ESS of above 200 for 1000 samples. We see that our 
distribution is quite well explored using independent random walk.

## b) Comparing prior and posterior

Let's sample from the prior distribution and posterior distribution
and plot both from in a colored scatterplot to see how our bayesian analysis
has updated our belief in $R$ and $S$.

```{r}
library(ggplot2)

#
# inverse transform algorithm
# (borrowed from mandatory 2 to sample R)
#
rltrunc_normal <- function(n,l,mu,sigma){
  Fl <- pnorm(q=l,mean=mu,sd=sigma)
  return(qnorm(p=(Fl + runif(n)*(1.0-Fl)),mean=mu,sd=sigma))
}

#
# simulate random variates from joint distribution
# (borrowed from mandatory to sample prior)
#
rRS <- function(n){
  R <- rltrunc_normal(n,1.8,2.2,1.5)
  S <- runif(n,min=0.5/R,0.8/R)
  return(cbind(R,S))
}

# Sample 1000 priors
m2 <- rRS(1000)

# Create vectors for dataframe
R <- c(m[,1], m2[,1])
S <- c(m[,2], m2[,2])
Distribution <- c(replicate(length(m[,1]), "Posterior"), 
                  replicate(length(m2[,1]), "Prior"))

# Create dataframe
df <- data.frame(R, S, Distribution)

ggplot(df, aes(R, S)) + geom_point(aes(colour = Distribution))
```

\newpage

# Bibliography