---
title: "Assignment 3"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: assignment2.bib
---

Clear R environment
```{r}
rm(list = ls())
```

# Problem 1 Random number generation and Monte Carlo integration.

## a) 

In this problem the task is to sample $n$ independent samples from the
probability distribution 

$$
p(x) = \frac{2^{\frac{1}{4}}\Gamma \left( \frac{3}{4}\right)}{\pi} \exp
  \left( -\frac{x^4}{2} \right) , \qquad -\infty < x < \infty 
$$

For this, we can use random walk metropolis hasting. We import the function 
from the lectures.

```{r}
# general 1d Gaussian proposal random walk MH
oneD.RWMH <- function(lprob, #notice log-density kernel!
                     sigma=1.0,
                     theta1=0.0,
                     n.iter=10000){
  # space for output
  output <- numeric(n.iter)
  # first iterate given
  output[1] <- theta1
  
  # main iteration loop
  for(t in 2:n.iter){
    # proposal
    thetaStar <- output[t-1] + rnorm(1,sd=sigma)
    # accept probability, for numerical stability we compute
    # the log-accept prob, and then take exp
    alpha <- exp(min(0.0,lprob(thetaStar)-lprob(output[t-1])))
    # accept/reject step
    if(runif(1)<alpha && is.finite(alpha)){
      output[t] <- thetaStar
    } else {
      output[t] <- output[t-1]
    }
  }
  return(output)
} 

# Logartihm of pdf (needed for metropolis hastings from slides)
logpdf <- function(x){
  c <- log((2^0.25 * gamma(3/4)) / pi)
  return(c - x^4/2)
}

# pdf from mandatory pdf
pdf <- function(x) {
  return ((2^0.25*gamma(3/4)/pi) * exp(-x^4 / 2))
}


# sample n samples
n <- 10000
out <- oneD.RWMH(logpdf, theta1 = 0.0, sigma=1, n.iter = n)

# plot 
par(mfrow=c(2,1))
plot(1:length(out),out,pch=20,cex=0.1,xlab="MCMC iteration #")
hist(out, probability = TRUE)
curve(pdf, add = TRUE, -2, 2)
```
We choose this algorithms since it is difficult to integrate the pdf to find the
cdf for inverse sampling, but the logarithm of the pdf is quite simple and 
metropolis hastings random walk is quite effective in low dimensions and for 
distributions that are difficult to sample directly.

From the histogram where we plot the pdf we see that the sampling is quite
effective. We also see from the plot that the random walk explores the 
distribution space quite nicely.

## b) 

Now let's sample from the distribution

$$
p(x) = 2x \exp(-x^2), \qquad 0 < x < \infty
$$
This function seems to be quite simple to find the CDF for and direct sampling 
via inverse transform sampling is possible. We get

$$
F(x) = \int_{0}^{x} p(x) = 1 - e^{-x^2}
$$
And by the inverse sampling method, we can calculate 

$$
X = F^{-1}(U) = \pm \sqrt{-\ln(1 - u)}
$$
Given the nature of calculating square roots, we get positive and negative 
X values but since the support of the pdf clearly states that $0 < x < \infty$
we reject the negative values.

in R:

```{R}
# Use inverse sampling method to sample from
# p(x) = 2x exp(-x^2)
inverse_sampling <- function(Nsamples) {
  # Sample n samples from uniform distribution
  u <- runif(Nsamples)
  # Return X = F^-1(U)
  return(sqrt(-log(1-u)))
}

# PDF
pdf2 <- function(x) {
  return(2*x * exp(-x^2))
}

x <- inverse_sampling(10000)
hist(x, probability = TRUE)
curve(pdf2, 0, 3, add = TRUE)
```

# c) 

Now we will consider the integral

$$
\int_0^\infty \exp(\sqrt{x})\exp(-20(x-4)^2) \, dx
$$
Let's evaluate the integral using importance sampling. First let's plot the 
function $g(x)$ that is being integrated:

In R:

```{r}
g <- function(x) {
 exp(sqrt(x))*exp(-20*(x-4)^2)
}

f <- function(x) {
  dnorm(x, mean=4, sd=0.1)
}

x<-seq(1,10,0.01); y1=g(x); y2=f(x)
plot(x, y1, type="l", pch=19, col="red", xlab="x", ylab="y")
lines(x, y2, pch=18, col="blue", type="l", lty=2)
legend(1, 95, legend=c("Line 1", "Line 2"),
       col=c("red", "blue"), lty=1:2, cex=0.8)
```
With the function we want to integrate in $g(x)$ in red and the $f(x)$ in 
dashed blue. We want to calculate $E\left(\frac{g(x)I(x \in A)}{f(x)}\right)$
by sampling $X$ values from $f$ and calculating the empirical mean of 
$\frac{g(x)I(x \in A)}{f(x)}$ where $I$ is the indicator function for the 
support.

```{r}
importance <- function(Nsamp) {
  # Sample from f
  x <- rnorm(Nsamp, 4, 0.1)
  # calculate empirical mean
  return(mean((g(x)*(x>0))/(f(x))))
}

# Estimate of integral using importance sampling
importance(100000)
# Numerical integration
integrate(g, 0, Inf)$value
```

# Problem 2. Smile shaped target

In this exercise, we shall sample from the distribution

$$
\log g(\boldsymbol \theta) = -\frac{\theta_1^2}{2} - 
\frac{(\theta_2 - \theta_1^2)^2}{2}, \qquad \infty < x < \infty
$$
To do this, an 2D random walk with multivariate normal proposals as been
implemented in R.

In the code below, a 2D random walk with proposal

$$
N(\boldsymbol \theta, \Sigma)
$$
Where 

$$
\Sigma = 
\begin{bmatrix}
2, 0 \\
0, 2
\end{bmatrix}
$$
and
$$
\boldsymbol \theta = 
\begin{pmatrix}
0 \\
0
\end{pmatrix}
$$
Yielded the greatest effective sample size.

In R:

```{r}
# Load the multivariate normal library 
library(mvtnorm)
library(coda)

# Target function from 
logg <- function(theta) {
  return(-theta[1]^2/2 - (theta[2] - theta[1]^2)^2 /2)
}

# 2D Random walk using log g
smile_shaped <- function(logg,
                         sigma=diag(2),
                         theta=c(0.0, 0.0),
                         Nsamp=100000){
  
  # Reserve 2xNsamp matrix with zeros
  res <- matrix(0, Nsamp, 2)
  
  # Set initial conditions
  res[1,] <- theta
  
  # Calculate old log-probability
  logold <- logg(theta)
  
  # accept counter
  Nacc <- 0
  
  # Iterate no of samples.
  for(i in 2:Nsamp){
    # Proposal step
    new <- res[i-1,] + rmvnorm(1, theta, sigma)
    # Log-p of proposal step
    lognew <- logg(new)
    # Evaluate step
    logfrac <- exp(min(0.0,lognew-logold))
    # Accept or reject new step
    if(runif(1)<logfrac && is.finite(logfrac)){
      # accept
      res[i,] <- new
      logold <- lognew
      Nacc <- Nacc+1
    } else {
      # reject
      res[i,] <- res[i-1,]
    }
  }
  print(paste("Accept prob: ", Nacc/Nsamp))
  return(res[5000:Nsamp,])
}

# Covariance matrix
sigma <- matrix(c(2, 0, 0, 2), nrow=2, ncol=2)
m <- smile_shaped(logg=logg, sigma=sigma, c(0.0, 0.0), 50000)
```

Let's plot the ``Smile shaped'' distribution:

```{r}
# Plot multivariate distribution
plot(m)
```

Let's also see if the marginal distribution of $\theta_1$ is standard normal:

```{r}
# Plot marginal of theta1 and compare with standard normal
hist(m[,1], probability=TRUE)
curve(dnorm(x, mean=0, sd=1), 
      col="red", lwd=2, add=TRUE, yaxt="n")
```

To check if we have explored the distribution well, let's calculate the effective
sample size of $theta_1$ and $theta_2$.

```{r}
ESS <- function(x) {
  coda::effectiveSize(x)
}

ESS(m[,1])
ESS(m[,2])
```
# Problem 3

```{r}
# Load the data set
df <- data.frame(read.table("logistic_regression_data.txt"))
x <- df$x
y <- df$y

# function returning a log-posterior kernel for theta
logistic.lp <- function(theta) {
  alpha <- theta[1]
  beta <- theta[2]
  # log-likelihood
  Eeta <- exp(alpha + beta*x)
  p <- Eeta/(1.0 + Eeta)
  log.like <- sum(dbinom(y, size=1, prob = p, log=TRUE))
  
  # priors
  log.prior <- dnorm(alpha, sd=10, log=TRUE) + dnorm(beta, sd=10, log=TRUE)
  
  # log-posterior kernel
  return(log.like+log.prior)
}
```

In this task, we will use independent metropolis-hastings sampling for the
posterior distribution $p(\boldsymbol \theta | \boldsymbol y)$ using a 
$N(\hat{\theta}, \delta \hat{\Sigma})$ proposal distribution.

In R:

```{r}

```
\newpage

# Bibliography