---
title: "Assignment 2"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
---

Clear R environment
```{r}
rm(list = ls())
```

# Problem 1

Consider the integral

$$
\int_{-1}^{1} \int_{-1}^{1} 1_D(x,y) dxdy
$$

Where $1_D(x, y)$ is the indicator function defined so that 
$$
1_D(x,y) = \begin{cases}
1 \quad \text{if} \quad x^2 + y^2 \leq 1, \newline
0 \quad \text{otherwise.}
\end{cases}
$$
As a crude first attempt, consider the estimator
$$
\theta_{CMC} = \frac{4}{N}\sum_{i=1}^{n} 1_D(X_i, Y_i)
$$
A.) Argue for why $\theta_{CMC}$ is a monte carlo estimator for the integral above.

According to [this](https://en.wikipedia.org/wiki/Monte_Carlo_integration#overview) 
wikipedia article which  was accessed the \formatdate{09}{10}{2020}, 
the monte carlo estimate for a multidimensional definite integral


$$
I = \int_\Omega f(\boldsymbol{\bar{x}}) d\boldsymbol{\bar{x}}
$$

where $\Omega$ is a subset of $R^m$, has volume


$$
V = \int_\Omega d\boldsymbol{\bar{x}}
$$

The naive Monte Carlo approach is to sample points uniformly on $\Omega$
given N uniform samples,

$$
\boldsymbol{\bar{x}_1}, \dots, \boldsymbol{\bar{x}_n} \in \Omega,
$$
I can be approximated by
$$
I \approx \Omega_N \equiv  V \frac{1}{N}\sum_{i=1}^{N} f(\boldsymbol{\bar{x}_i})
$$

Which is true due to the law of large numbers.

In our case, which is also very similar to the 
[example](https://en.wikipedia.org/wiki/Monte_Carlo_integration#Example) on the 
wikipedia article for monte carlo integration, $\Omega = [-1, 1] \times [-1, 1]$
with $V = \int_{-1}^1 \int_{-1}^1 dxdy = 4$ which gives the following crude way 
to estimate $I$
$$
I = \frac{4}{N}\sum_{i=1}^{N}1_D(X_i, Y_i) 
$$

Which is the proposed estimator $\theta_{CMC}$.

A.) Show that $1_D(X_i, Y_i)$ has a bernoulli distribution with $p=\frac{\pi}{4}$

The function returns success or failure, and is therefore has aa potential '
bernoulli distribtion. To calculate $P(X^2 + Y^2 \leq 1)$ we need to calculate
the 

c) Implementation of monte carlo estimate of $\theta_{CMC}$ with $N = 1000$
 
```{r}
indicator1 <- function (x, y) {          # Indicator function for unit circle
  return((x^2+y^2 <= 1))
}

mcEstimatePi <- function (Sims) {     # function for monte carlo estimate 
  mcPi <- numeric(Sims)
  for (i in 1:Sims) {
    N <- 1000                            # Number of samples
    x <- runif(N, -1, 1)                 # x values from uniform
    y <- runif(N, -1, 1)                 # y values from uniform
    mcPi[i] <- (4/N)*sum(indicator1(x, y)) # return monte carlo estimate
  }
  return(mcPi)
}

Sims <- 10000                            # Simulations of pi
mcPi <- mcEstimatePi(Sims)        # Function call
summary(mcPi)                            # Summary stats for MC estimate
hist(mcPi)                               # Histogram
```
```{r}
mean(mcPi)
var(mcPi)
```

We see that the distribution is approxamately normally distrobuted with sample
mean at approx $Mean = 3.141$ and sample variance at approx $Variance =  0.00260$.
This is du to the central limit theorem because our estimate of pi is based on a
sum of several samples.

d.) Calculate probability of correctly estimating to two decimal places

```{r}
mean(mcPi < 3.15) - mean(mcPi <= 3.14)
```
e.) Introducing antithetic variables

The assignment proposes two antithetic variables $V = a + b - X = -X$ and 
$W = -Y$

These will not reduce the monte carlo variance since $X$ and $Y$ are independent
uniform random variables and due to their independence there is no negative
correlation gained by just flipping the sign of both variables from $X$ and $Y$ to  
$-X$ and $-Y$.
```{r}
x <- runif(1000, -1, 1)
y <- runif(1000, -1, 1)
cov(-x, -y)
```

f.) Let's see if our variance is reduced by doing exercise c. with the new variables.

```{r}
mcEstimatePi <- function (Sims) {     # function for monte carlo estimate 
  mcPi <- numeric(Sims)
  for (i in 1:Sims) {
    N <- 1000                            # Number of samples
    x <- runif(N, -1, 1)                 # x values from uniform
    y <- runif(N, -1, 1)                 # y values from uniform
    mcPi[i] <- (4/N)*sum(indicator1(-x, -y)) # mcEstimate with new variables
  }
  return(mcPi)
}

Sims <- 10000                            # Simulations of pi
mcPi <- mcEstimatePi(Sims)               # Function call
summary(mcPi)                            # Summary stats for MC estimate
mean(mcPi)
var(mcPi)
```
g.) Check if shift function reduces variance

```{r}
shift <- function(u) {                   # Introducing shift function
  return(((u+2.0) %% 2.0) - 1.0)
}

mcEstimatePi <- function (Sims) {     # function for monte carlo estimate 
  mcPi <- numeric(Sims)
  for (i in 1:Sims) {
    N <- 1000                            # Number of samples
    x <- runif(N, -1, 1)                 # x values from uniform
    y <- runif(N, -1, 1)                 # y values from uniform
    sx <- shift(x)
    sy <- shift(y)
    mcPi[i] <- (4/N)*sum(indicator1(sx, sy)) # mcEstimate with new variables
  }
  return(mcPi)
}

Sims <- 10000                            # Simulations of pi
mcPi <- mcEstimatePi(Sims)               # Function call
summary(mcPi)                            # Summary stats for MC estimate
mean(mcPi)
var(mcPi)
```
h.) Use important sampling

```{r}
f <- function (x, y, sigma) {
  return((1/(2*pi*sigma^2))*exp(-(x^2)/(2*sigma^2))*exp(-(y^2)/(2*sigma^2)))
}

mcEstimatePi <- function (Sims, sigma) {  # function for monte carlo estimate 
  mcPi <- numeric(Sims)
  for (i in 1:Sims) {
    N <- 1000                                 # Number of samples
    x <- rnorm(N, 0, sigma^2)                 # x values from uniform
    y <- rnorm(N, 0, sigma^2)                 # y values from uniform
    mcPi[i] <- mean(indicator1(x, y) / f(x, y, sigma))
  }
  return(mcPi)
}
```

```{r}
mcPi <- mcEstimatePi(10000, 0.3)
summary(mcPi)
```

# Problem 2

Define $\lambda(t)$ in R 
```{r}
# Specify the intensity function for storms
lambdastorm <- function(t) {
 (297 / 10)*(1 + cos(2*pi*(t + (1/10)))) * (1 - (exp(-t/10)/2)) + 3/5
}

lambdastorm(0.5)
```

a.) Calculating number of storms, expected value and variance

According to Rizzo on page 103. A poisson proccess with a intensity function
$\lambda(t)$ has the property that the number of events $N(t)$ in interval $[0, t]$ 
has the poisson distribution with mean 
$$
E[N(t)] = \int_0^t \lambda(y) dy
$$
Which in our case gives 

```{r}
integrate(lambdastorm, 0, 1)
```
Which we have confirmed by the simulation above which has a simulated mean of 
approximately $16.3$

Let's find the expected number of storms in 2025 by calculating the integral
```{r}
integrate(lambdastorm, 5, 6)
```
And let's find the expected value and standard deviation of storms in 
2020 and 2021 combined

Expected value is calculated using the integral below
```{r}
integrate(lambdastorm, 0, 2)
```

Which means the number of events in 2020 and 2021 combined is poisson
distrubuted with $\lambda = 39.72776$. The variance of a poisson distribution is
$Var(X) = \lambda$ (according to Rizzo on page 44).  
$$
SD(X) = \sqrt{Var(X)} = \sqrt{33.92776 } =  5.824754
$$
b.) Find smallest possible $\lambda_{max}$ for all $\lambda(t), \quad t \geq 0$

The function $\lambda(t)$ does not have a global maximum because it is modeled
with a increasing winter intensity due to climate change that does not stop
increasing. Solving for $\frac{d}{dt} \lambda(t) = 0$ gives an infinite number
of potential maximum or minimum points and there is no $\lambda_{max}$ for all
$\lambda(t)$ values.

c.) Validate previous points by simulation

simtNHPP borrowed from lectures on stochastic processes
```{r}
# Function for simulating arrival times for a NHPP between a and b using thinning
simtNHPP <- function(a,b,lambdamax,lambdafunc){
  # Simple check that a not too small lambdamax is set
  if(max(lambdafunc(seq(a,b,length.out = 100)))>lambdamax)
    stop("lambdamax is smaller than max of the lambdafunction")
  # First simulate HPP with intensity lambdamax on a to b
  expectednumber <- (b-a)*lambdamax  
  Nsim <- 3*expectednumber  # Simulate more than the expected number to be certain to exceed stoptime
  timesbetween <- rexp(Nsim,lambdamax) # Simulate interarrival times
  timesto <- a+cumsum(timesbetween)   # Calculate arrival times starting at a
  timesto <- timesto[timesto<b] # Dischard the times larger than b
  Nevents <- length(timesto) # Count the number of events
  # Next do the thinning. Only keep the times where u<lambda(s)/lambdamax
  U <- runif(Nevents)
  timesto <- timesto[U<lambdafunc(timesto)/lambdamax]  
  timesto  # Return the remaining times
}

Nsim <- 1000
a <- 0
b <- 1
NHPPnumbers <- vector(length=Nsim)
for(i in 1:Nsim) {
  NHPPnumbers[i] <- length(simtNHPP(a=a,b=b,
                           lambdamax=max(lambdastorm(seq(a, b, 0.01))),
                           lambdafunc=lambdastorm))
}

# Exepcted number of storms in 2020
mean(NHPPnumbers)

Nsim <- 1000
a <- 5
b <- 6
NHPPnumbers <- vector(length=Nsim)
for(i in 1:Nsim) {
  NHPPnumbers[i] <- length(simtNHPP(a=a, b=b,
                           lambdamax=max(lambdastorm(seq(a, b, 0.01))),
                           lambdafunc=lambdastorm))
}

# Exepcted number of storms in 2025
mean(NHPPnumbers)

Nsim <- 1000
a <- 0
b <- 2
NHPPnumbers <- vector(length=Nsim)
for(i in 1:Nsim) {
  NHPPnumbers[i] <- length(simtNHPP(a=a, b=b,
                           lambdamax=max(lambdastorm(seq(a, b, 0.01))),
                           lambdafunc=lambdastorm))
}

# Expected number of storms in 2020 and 2021
mean(NHPPnumbers)
# Variance of number of storms in 2020 and 2021
var(NHPPnumbers)
# Standard deviation of number of storms in 2020 and 2021
sd(NHPPnumbers)
```
