---
title: "Assignment 1"
author: "Bjørn Christian Weinbach"
date: "18.09.2020"
header-includes:
   - \usepackage{bbm}
output: pdf_document
---

# Problem 1

## a.) Verify that this is a proper probability density function.

We have the proposed pdf

\begin{equation}\label{pdfprob1}
f(x) = \frac{1}{2} - \frac{x}{4}, \quad -1 \leq x \leq 1
\end{equation}

which is 0 whenever $x \notin [-1, 1]$

To be qualified as a pdf the function need to satisfy the following:

1.
\begin{equation}
  \int_{-\infty}^{\infty} f(x) dx = 1
\end{equation}

2.
\begin{equation}
  f(x) \geq 0, \qquad \text{for any x}
\end{equation}

Condition 2. can easily be validated by inspection. any x in the interval
-1 to 1 will yield a positive value.

The condition 1, must be validated through integration. We do this analytically
like so

\begin{equation}
  \int_{-\infty}^{\infty} f(x) dx = \int_{-1}^{1} \frac{1}{2} - \frac{x}{4} dx
  = \left[\frac{x}{2} - \frac{x^2}{4}\right]_{-1}^{1} = 1
\end{equation}

## a.) Find the mean $E(X)$ and the variance $Var(x)$.

Expectation is formally defined as 

\begin{equation}
E[X] = \int_{\mathbb{R}} x f(x) dx
\end{equation}

We do this analytically and get

\begin{equation}
E[X] = \int_{\mathbb{R}} x f(x) dx = \int_{-1}^{1} \frac{x}{2} - \frac{x^2}{4} dx
= \left[\frac{x^2}{4} - \frac{x^3}{12}\right]_{-1}^{1} = -\frac{1}{6}
\end{equation}

Variance is defined as

\begin{equation}
Var(X) = \int_{\mathbb{R}} x^2 f(x) \,dx - \mu^2 
\end{equation}

Where $\mu = E[X]$

We do this analytically like so

\begin{equation}
Var(X) = \int_{\mathbb{R}} x^2 f(x) \,dx - \mu^2  = 
\int_{-1}^{1} \frac{x^2}{2} - \frac{x^3}{4} dx - (-\frac{1}{6})^2 =
\left[\frac{x^3}{12} - \frac{x^4}{16}\right]_{-1}^{1} - \frac{1}{36} = \frac{11}{36}
\end{equation}

## b.) Find the cumulative distribution function $F(x)$ associated with $X$, 
and use this to calculate $P(X < 0)$ and $P(-0.5 < X < 0.5)$. 

$F_{X}(x)$ is formally defined as

\begin{equation}
  F_X(x) = \int_{-\infty}^x f_X(t) dt 
\end{equation}

So in our case, the cumulative distribution function for f(x) is

\begin{equation}
F_X(x) = \int_{-\infty}^x f_X(t)\,dt =
\int_{-1}^{x} \frac{1}{2} - \frac{x}{4} dt =
\left[\frac{t}{2} - \frac{t^2}{4} \right]_{-1}^{x} =
\frac{1}{8}(-x^2 + 4x +5)
\end{equation}

Since $F_X(x) = P(X < x)$, we can calculate the probabilities above.
This gives us that $P(X < 0) = \frac{5}{8} = 0.652$ and $P(-0.5 < X < 0.5) = 
P(X < 0.5) - P(X < -0.5) = 0.84375 - 0.34375 = 0.5$ 

To find $F^{-1}_X(u)$ we set $U = \frac{1}{8}(-x^2 + 4x +5)$ and solve for x
which gives us

\begin{equation}
  X = 2 \pm \sqrt{9 - 8u}
\end{equation}

And we choose $X = 2 - \sqrt{9 - 8u}$ since we will use this together with
the inverse transformation method to generate numbers $X$ from the distribution
above and this particular solution is the only one that will return values in 
the inverval $[-1, 1]$

c.) Write a function (with single argument n) which produces n random numbers
with density (\ref{pdfprob1})

```{r}
# itm (Inverse Transform Method) function for equation one in assignment 1
itm_equation1 <- function(n) {
  u <- runif(n)
  return (2 - sqrt(9 - 8*u))
}
```

Check that your function produces correct results by comparing a histogram

```{r}
Nsim <- 1000000
itm_p <- itm_equation1(Nsim)

hist(itm_p, breaks = 30, prob = TRUE, xlab = "X", ylab = "P(X)", 
     main="Histrogram of inverse transform method")
lines(density(itm_p))
```

And to see that this is correct, we plot the pdf provided by assignment 1
to see that this indeed generate numbers from the given distribution.

```{r}
theoretical_pdf <- function(x) {
  return((1/2) - (x/4))
}

x <- seq(-1, 1, 0.01)
plot(x, theoretical_pdf(x), main = "Plot of pdf", xlab = "x", ylab = "f(x)", 
     type ="l")
```

d.) Using a $uniform(-1,1)$ proposal distribution $g$, do the calculations required 
to obtain an accept-reject method for drawing random numbers with 
density (\ref{pdfprob1})

The uniform distribution in interval $[a, b]$ has pdf
\begin{equation}
f_{u(a, b)}(x) = \frac{1}{b-a}, \qquad a \leq x \leq b
\end{equation}

Which we will use later. The algorithm for acceptance-rejection algorithm is as
follows:

1. generate y from g

where g is the proposal, in our case $uniform(-1, 1)$.

2. generate $U$ from $uniform(0, 1)$

3. accept y if $u < \frac{f(y)}{c g(y)}$ and let $x = y$

to find the constant c. such that $\frac{f(t)}{g(t)} \leq c$ for all $t$ where
$f(t) > 0$. In our case, this means finding the maximum value of the equation
\begin{equation}
\frac{f(t)}{g(t)} = \frac{f_X(x)}{f_{u(-1,1)}(x)} = 
\frac{\frac{1}{2} - \frac{x}{4}}{\frac{1}{2}}  \implies 1 - \frac{x}{2},
\qquad -1 \leq x \leq 1
\end{equation}

Which has maximum value at $x = -1$ which gives $f(-1) = 1 + \frac{1}{2} = 1.5$
which means our constant $c = 1.5$.


e.) Write a function (with single argument n) which produces n random numbers 
with density (\ref{pdfprob1}) using the accept-reject method. Check the results 
in the same way as in point c.Which method would you prefer in this case - 
accept-reject or inverse transform? The function system.time is useful in this 
regard.

We apply the algorithm described above in R:

```{r}
# ar (acceptance rejection) for equation 1
ar_equation1 <- function(n) {
  f <- function(x) (1/2 - x/4)
  C = 1.5
  
   naccepts <- 0
   result.sample <- rep(NA, n)
   
   while (naccepts < n) {
    y <- runif(1, -1, 1)
    u <- runif(1)
   
    if ( u <= f(y) / (C*(1/2))) {
      naccepts <- naccepts + 1
      result.sample[naccepts] = y
    }
   }
   
   result.sample
}

Nsim <- 1000000
ar_p <- ar_equation1(Nsim)
hist(ar_p, probability = TRUE, main="Histogram of acceptance rejection method",
     xlab = "X", ylab = "P(X)", breaks=30)
lines(density(ar_p))
```

This function is alot slower than the inverse transform method, mainly because
we iterate until n samples have been generated instead of using vectorized code.
This slows down the code because of R being an interpereted language. 

You can write a vectorized version which takes into account the probability
of accepting a propasal and scaling up the number of simulations accordingly, 
but this does not guarantee that N samples is generated since the expected
number of iterations needed is a geometric distrubution with expected value
$E[X] = \frac{1}{p(accept)}$

f.) Check the answers you got for expectations, variance and probabilities in
points a) and b) above by simulation.

Let's begin with expectation. Analytically, the expectation is $E[X] = 
-\frac{1}{6}$. In R, the simulated smaple is:

```{r}
# Mean of inverse transform method simulated values
mean(itm_p)
#Mean of acceptance rejection simulated values
mean(ar_p)
# Analytical mean
-1/6
```

We se that both methods generate the correct mean that we have calculated.

Now let's check the variance:

```{r}
# Variance of inverse transform method simulated values
var(itm_p)
# Variance of acceptance rejection simulated values
var(ar_p)
# Analytical variance
11/36
```

To find the 95\% confidence interval, we first take 100 samples and calculate
the mean and do this for every 100 random sample in our data (this gives us 
10000 sample means).

```{r}
m <- matrix(itm_p, Nsim/100, 100)
sample_means <- rowMeans(m)
hist(sample_means, breaks = 30)
```

And we see that the sample means are normally distributed, which is confirmed by
the central limit theorem. We can then use the formula for 
confidence intervals using Z-scores

\begin{equation}
CI = \bar{X}\pm Z \frac{\sigma}{\sqrt{n}}
\end{equation}

In R, we get:

```{r}
CI <- function(data) {
  xhat <- mean(data)
  sd <- sd(data)
  n <- length(data)
  lower <- xhat - 2*(sd / n)
  upper <- xhat + 2*(sd / n)
  return(c(lower, upper))
}

CI(sample_means)
```

# Problem 2