---
title: "Assignment 3"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: assignment3.bib
---

Clear R environment
```{r}
rm(list = ls())
```

# Problem 1. Random number generation and Monte Carlo integration.

## a) 

In this problem the task is to sample $n$ independent samples from the
probability distribution 

$$
p(x) = \frac{2^{\frac{1}{4}}\Gamma \left( \frac{3}{4}\right)}{\pi} \exp
  \left( -\frac{x^4}{2} \right) , \qquad -\infty < x < \infty 
$$

One might wish to implement a regular Metropolis-Hastings random walk for
sampling this distribution, but there is one major disadvantage for using this 
approach: [@wiki:metropolisco]

The samples are correlated, even though over the long term they do correctly 
follow $p(x)$. This is due to the fact that given that you know the current
time step, you have a pretty good estimate of what the value of the next one is.
In short, the adjacent samples are autocorrelated.

Because of this and the task specifically asking for independent samples.
A regular random walk metropolis-hastings approach is not feasible, we then turn
to the independent metropolis-hastings approach:

In R:

```{r}
# general 1d independence metropolis hastings
oneD.IRWMH <- function(prob,
                       sigma=1.0,
                       theta1=0.0,
                       Nsamp=10000){
  res <- numeric(Nsamp)
  # allocate memory
  res[1] <- theta1
  # old importance weight
  wt.old <- prob(res[1])/dnorm(res[1], sd=sigma)
  Nacc <- 0
    for(i in 2:Nsamp){
      # proposal (note, independent of past)
      thetaStar <- rnorm(1, sd=sigma)
      # new importance weight
      wt.star <- prob(thetaStar)/dnorm(thetaStar, sd=sigma)
      # accept probability
      alpha <- min(1.0, wt.star/wt.old)
      # accept/reject
      if(runif(1) < alpha){
        res[i] <- thetaStar
        wt.old <- wt.star
        Nacc <- Nacc+1
      } else {
        res[i] <- res[i-1]
      }
    }
  return(res)
}

# pdf from mandatory pdf
pdf <- function(x) {
  return ((2^0.25*gamma(3/4)/pi) * exp(-x^4 / 2))
}


# sample n samples
n <- 10000
out <- oneD.IRWMH(pdf, theta1 = 0.0, sigma=1, Nsamp = n)

# plot 
par(mfrow=c(2,1))
plot(1:length(out),out,pch=20,cex=0.1,xlab="MCMC iteration #")
hist(out, probability = TRUE)
curve(pdf, add = TRUE, -2, 2)
```
We choose this algorithms since it is difficult to integrate the pdf to find the
cdf for inverse sampling, but the logarithm of the pdf is quite simple and 
metropolis hastings random walk is quite effective in low dimensions and for 
distributions that are difficult to sample directly.

From the histogram where we plot the pdf we see that the sampling is quite
effective. We also see from the plot that the random walk explores the 
distribution space quite nicely.

## b) 

Now let's sample from the distribution

$$
p(x) = 2x \exp(-x^2), \qquad 0 < x < \infty
$$
This function seems to be quite simple to find the CDF for and direct sampling 
via inverse transform sampling is possible. We get

$$
F(x) = \int_{0}^{x} p(x) = 1 - e^{-x^2}
$$
And by the inverse sampling method, we can calculate 

$$
X = F^{-1}(U) = \pm \sqrt{-\ln(1 - u)}
$$
Given the nature of calculating square roots, we get positive and negative 
X values but since the support of the pdf clearly states that $0 < x < \infty$
we reject the negative values.

in R:

```{R}
# Use inverse sampling method to sample from
# p(x) = 2x exp(-x^2)
inverse_sampling <- function(Nsamples) {
  # Sample n samples from uniform distribution
  u <- runif(Nsamples)
  # Return X = F^-1(U)
  return(sqrt(-log(1-u)))
}

# PDF
pdf2 <- function(x) {
  return(2*x * exp(-x^2))
}

x <- inverse_sampling(10000)
hist(x, probability = TRUE)
curve(pdf2, 0, 3, add = TRUE)
```

## c) 

Now we will consider the integral

$$
\int_0^\infty \exp(\sqrt{x})\exp(-20(x-4)^2) \, dx
$$
Let's evaluate the integral using importance sampling. First let's plot the 
function $g(x)$ that is being integrated:

In R:

```{r}
g <- function(x) {
 exp(sqrt(x))*exp(-20*(x-4)^2)
}

f <- function(x) {
  dnorm(x, mean=4, sd=0.1)
}

x<-seq(1,10,0.01); y1=g(x); y2=f(x)
plot(x, y1, type="l", pch=19, col="red", xlab="x", ylab="y")
lines(x, y2, pch=18, col="blue", type="l", lty=2)
legend(1, 95, legend=c("Line 1", "Line 2"),
       col=c("red", "blue"), lty=1:2, cex=0.8)
```
With the function we want to integrate in $g(x)$ in red and the $f(x)$ in 
dashed blue. We want to calculate $E\left(\frac{g(x)I(x \in A)}{f(x)}\right)$
by sampling $X$ values from $f$ and calculating the empirical mean of 
$\frac{g(x)I(x \in A)}{f(x)}$ where $I$ is the indicator function for the 
support.

```{r}
importance <- function(Nsamp) {
  # Sample from f
  x <- rnorm(Nsamp, 4, 0.1)
  # calculate empirical mean
  return(mean((g(x)*(x>0))/(f(x))))
}

# Estimate of integral using importance sampling
importance(100000)
# Numerical integration
integrate(g, 0, Inf)$value
```

# Problem 2. Smile shaped target

## a)

In this exercise, we shall sample from the distribution

$$
\log g(\boldsymbol \theta) = -\frac{\theta_1^2}{2} - 
\frac{(\theta_2 - \theta_1^2)^2}{2}, \qquad \infty < x < \infty
$$
To do this, an 2D random walk with multivariate normal proposals as been
implemented in R.

In the code below, a 2D random walk with proposal

$$
N(\boldsymbol \theta, \Sigma)
$$
Where 

$$
\Sigma = 
\begin{bmatrix}
2, 0 \\
0, 2
\end{bmatrix}
$$
and
$$
\boldsymbol \theta = 
\begin{pmatrix}
0 \\
0
\end{pmatrix}
$$
Yielded the greatest effective sample size.

In R:

```{r}
# Load the multivariate normal library 
library(mvtnorm)
library(coda)

# Target function from 
logg <- function(theta) {
  return(-theta[1]^2/2 - (theta[2] - theta[1]^2)^2 /2)
}

# 2D Random walk using log g
smile_shaped <- function(logg,
                         sigma=diag(2),
                         theta=c(0.0, 0.0),
                         Nsamp=100000){
  
  # Reserve 2xNsamp matrix with zeros
  res <- matrix(0, Nsamp, 2)
  
  # Set initial conditions
  res[1,] <- theta
  
  # Calculate old log-probability
  logold <- logg(theta)
  
  # accept counter
  Nacc <- 0
  
  # Iterate no of samples.
  for(i in 2:Nsamp){
    # Proposal step
    new <- res[i-1,] + rmvnorm(1, theta, sigma)
    # Log-p of proposal step
    lognew <- logg(new)
    # Evaluate step
    logfrac <- exp(min(0.0,lognew-logold))
    # Accept or reject new step
    if(runif(1)<logfrac && is.finite(logfrac)){
      # accept
      res[i,] <- new
      logold <- lognew
      Nacc <- Nacc+1
    } else {
      # reject
      res[i,] <- res[i-1,]
    }
  }
  print(paste("Accept prob: ", Nacc/Nsamp))
  return(res[5000:Nsamp,])
}

# Covariance matrix
sigma <- matrix(c(2, 0, 0, 2), nrow=2, ncol=2)
m <- smile_shaped(logg=logg, sigma=sigma, c(0.0, 0.0), 50000)
```

Let's plot the ``Smile shaped'' distribution:

```{r}
# Plot multivariate distribution
plot(m)
```

Let's also see if the marginal distribution of $\theta_1$ is standard normal:

```{r}
# Plot marginal of theta1 and compare with standard normal
hist(m[,1], probability=TRUE)
curve(dnorm(x, mean=0, sd=1), 
      col="red", lwd=2, add=TRUE, yaxt="n")
```

To check if we have explored the distribution well, let's calculate the effective
sample size of $theta_1$ and $theta_2$.

```{r}
ESS <- function(x) {
  coda::effectiveSize(x)
}

ESS(m[,1])
ESS(m[,2])
```
# Problem 3. IMH for simple logistic regression problem

```{r}
# Load the data set
df <- data.frame(read.table("logistic_regression_data.txt"))
x <- df$x
y <- df$y

# function returning a log-posterior kernel for theta
logistic.lp <- function(theta) {
  alpha <- theta[1]
  beta <- theta[2]
  # log-likelihood
  Eeta <- exp(alpha + beta*x)
  p <- Eeta/(1.0 + Eeta)
  log.like <- sum(dbinom(y, size=1, prob = p, log=TRUE))
  
  # priors
  log.prior <- dnorm(alpha, sd=10, log=TRUE) + dnorm(beta, sd=10, log=TRUE)
  
  # log-posterior kernel
  return(log.like+log.prior)
}
```

## a) Sample $\alpha$ and $\beta$ using metropolis-hasting and bayesian inference

In this task, we will use independent metropolis-hastings sampling for the
posterior distribution $p(\boldsymbol \theta | \boldsymbol y)$ using a 
$N(\hat{\theta}, \delta \hat{\Sigma})$ proposal distribution.

In R:

```{r}
# Independent sampler
IndepMH <- function(lprob, theta=c(0, 0), sigma=diag(2), Nsamp=1000, S=1){
  # Allocate space
  res <- matrix(0, Nsamp, 2)
  # Set initial values
  res[1,] <- theta
  # Old importance
  pold <- lprob(theta) - dmvnorm(theta, mean=theta ,sigma=sigma, log=TRUE)
  # No of accept
  Nacc <- 0
  for (i in 2:Nsamp){
    # Sample theta from multivariate normal with scaled sigma
    new <- rmvnorm(1, theta, S*sigma)
    # Importance of new theta
    pnew <- lprob(new) - dmvnorm(new, mean=theta, sigma=sigma, log=TRUE)
    frac <- exp(min(0.0, pnew-pold))
    if(runif(1)<frac && is.finite(frac)){
      # Accept
      res[i,] <- new
      pold <- pnew
      Nacc <- Nacc+1
    } else {
      # Reject
      res[i,] <- res[(i-1),]
    }
  }
  print(paste0("accept rate : ",Nacc/Nsamp))
  return(res[1000:Nsamp,])
}

sigma <- matrix(c(0.00653,-0.00058,-0.00058,0.01689),2,2)

m <- IndepMH(logistic.lp, c(-0.102, 1.993), sigma, 10000, 0.6)
plot(m)
```

Calculate ESS

```{r}
ESS(m[,1])
ESS(m[,2])
```

b/c) Plot quantiles of data and find $x*$ such that $P(m(x*) > 0.8) = 0.99$

To  solve this problem, we calculate $m(x*)$ the median and quantiles for
$x \in [-5, 5]$ where $\alpha$ and $\beta$ has been estimated based on the
dataset provided in the mandatory assignment.

We use the following algorithm to solve this problem:

1. Given data, estimate $\alpha$, $\beta$ using independent metropolis hastings
2. Select sequence of $x*$ we want to explore
3. Iterate through each value of $x*$
4. Calculate median and 01th, 05th and 95th quantiles and return these as a vector
5. Store vector as row in matrix
6. Repeat steps 4-5 for each value in $x*$

When this algorithm terminates, we have a 2D matrix of medians and quantiles for
all $x*$. We use this for plotting and finding the $x*$ where 
$P(m(x*) > 0.8) = 0.99$

In R:

```{r}
# Plotdist is function for calculating values for plotting
plotdist <- function (x) {
  # Calculate m(x*)
  mstar <- exp(m[,1] + m[,2]*x) / (1 + exp(m[,1] + m[,2]*x))
  # calulcate median given x*
  med <- median(mstar)
  # Caltulate quantiles given x*
  quant05 <- quantile(mstar, 0.05)
  quant95 <- quantile(mstar, 0.95)
  quant01 <- quantile(mstar, 0.01)
  # return vector of calculations
  return(c(med, quant05, quant95, quant01))
}

# sequence of x* values
x <- seq(-5, 5, 0.01)
# plot 
plotmat <- matrix(0, length(x), 4)
# set index
i <- 0
# Loop through all x* values in sequence
for (val in x) {
  # increment index for each value in xÆ
  i <- i+1
  plotmat[i,] <- plotdist(val)
}

# First value larger than 0.8 (for plotting horizontal)
horizontal <- plotmat[, 4][plotmat[, 4] > 0.8][1]

# First x-value of m-value with 99% probability of being larger than 0.8
vertical <- x[Position(function(x) x > 0.8, plotmat[, 4])]

# Plot all medians and quantiles conditional on x*
plot(x, plotmat[, 1], type="l", col="red", ylab="M*")
lines(x, plotmat[, 2], type="l", col="blue", lty=2)
lines(x, plotmat[, 3], type="l", col="orange", lty=3)
lines(x, plotmat[, 4], type="l", col="green", lty=4)
# Add vertical line for x values
abline(v = vertical, col="black")
# add horizontal line for m values
abline(h = horizontal, col="black")
legend("topleft", legend=c("Median", "05 quantile", "95 quantile", "01 quantile"),
       col=c("red", "blue", "orange", "green"), lty=1:4, cex=0.8)

# X* where P(M(X*) > 0.8) = 0.99 
print(vertical)
```

# Problem 4. Gibbs sampler for simple linear regression model

## a) Find conditional posteriors.

The posterior for $\tau$ is given as

$$
\pi(\tau | \alpha, \beta, \boldsymbol{y}) \sim
\text{gamma}\left(\frac{n}{2} + 1, \frac{1}{\frac{1}{2}\sum_{i=1}^n
(y_i -\alpha - \beta x_i)^2 + 1
}\right)
$$

To calculate the posterior of $\alpha$ we see by inspection that the log-density
kernel is possibly on the form 
$\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) = a + b\alpha + c\alpha^2$ which 
tells us the distribution is normal and that the constant $a$ is unimportant.

Let's calculate $\log \pi(\alpha | \beta, \tau, \boldsymbol y)$.

From the log-density kernel in the assignment description, we get

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c'_\alpha - \frac{\tau}{2}\sum_{i=1}^n (y_i - \alpha - \beta x_i)^2 - 
\frac{\alpha^2}{200}.
$$
Where $c'_\alpha$ is a unimportant constant containing all values not dependent 
on $\alpha$. By squaring the term in the sum and adding unimportant values to
$c'_\alpha$ we now get.

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c'_\alpha - \frac{\tau}{2}(n \alpha^2 + \alpha \sum_{i=1}^n(2\beta x_i - 2y_i)) - 
\frac{\alpha^2}{200}
$$
By multiplying the $\tau$ term and reshuffling we get

$$
\log \pi(\alpha | \beta, \tau, \boldsymbol{y}) =
c_\alpha - \tau\sum_{i=1}^n(\beta x_i - y_i)\alpha -\left(\frac{\tau n}{2} + 
\frac{1}{200}\right)\alpha^2
$$
And we have

$$
b_\alpha = -\tau\sum_{i=1}^n(\beta x_i - y_i)
$$
and

$$
c_\alpha = -\left(\frac{\tau n}{2} + 
\frac{1}{200}\right)
$$

and the posterior (from the lecture notes) is

$$
\pi(\alpha | \beta, \tau, \boldsymbol{y}) \sim
N\left( -\frac{b_\alpha}{2c_\alpha},  -\frac{1}{2c_\alpha} \right)
$$

And we do the same for $\pi(\beta | \alpha, \tau, \boldsymbol{y})$

From the log-density kernel in the assignment description, we get

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \frac{\tau}{2}\sum_{i=1}^n (y_i - \alpha - \beta x_i)^2 - 
\frac{\beta^2}{200}.
$$
Where $c'_\beta$ is a unimportant constant containing all values not dependent 
on $\beta$. By squaring the term in the sum and adding unimportant values to 
$c'_\beta$ we now get.

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \frac{\tau}{2}\sum_{i=1 }^n \left(\beta^2 x_i^2 + 
2\alpha\beta x_i - 2\beta x_i y_i
\right) - \frac{\beta^2}{200}
$$
By multiplying the $\tau$ term and reshuffling we get

$$
\log \pi(\beta | \alpha, \tau, \boldsymbol{y}) =
c'_\beta - \tau\left(\sum_{i=1}^n \alpha x_i + \sum_{i=1}^n x_i y_i \right) -
\left(\frac{\tau}{2}\sum_{i=1}^n x_i + \frac{1}{200}\right)\beta^2
$$
And we have

$$
b_\beta = 
-\tau\left(\sum_{i=1}^n \alpha x_i + \sum_{i=1}^n x_i y_i \right)
$$
and

$$
c_\beta = 
-\left(\frac{\tau}{2}\sum_{i=1}^n x_i + \frac{1}{200}\right)
$$

and the posterior (from the lecture notes) is

$$
\pi(\beta | \alpha, \tau, \boldsymbol{y}) \sim
N\left( -\frac{b_\beta}{2c_\beta},  -\frac{1}{2c_\beta} \right)
$$

## b) Gibbs sampler with three blocks

Load the data

```{r}
# Load regression data
df <- data.frame(read.table(file="linear_regression_data.txt"))
x <- df$x
y <- df$y
``` 

Let's implement the gibbs sampler

```{r}
gibbs <- function(x, y, theta=c(1.0, 1.0, 1.0), Nsamp=10000) {
  res <- matrix(0.0, Nsamp, 3)
  res[1, ] <- theta
  gammashape <- length(x)/2 + 1
  
  for(i in 1:Nsamp){
    # Block 1. Tau.
    gammascale <- 0.5*(sum(x) - length(x)*res[i, 2] - sum(res[i, 3]*x))^2 + 1
    res[i, 1] <- rgamma(1, gammashape, gammascale)
    
    # Block 2. Alpha.
    balpha <- -res[i, 1]*sum(res[i, 1]*x - y)
    calpha <- -(res[i, 1]*length(x))/2 + (1/200)
    res[i, 2] <- rnorm(1, -(balpha/(2*calpha)), -(1/(2*calpha)))
    
        
    # Block 2. Beta
    bbeta <- -res[i, 1]*(res[i, 2]*sum(x) + sum(x*y))
    cbeta <- -((res[i, 1]/2)*sum(x) + 1/200)
    res[i, 3] <- rnorm(1, -(bbeta/(2*cbeta)), -(1/(2*cbeta)))
  }
  return(res)
}

m <- gibbs(x, y)
```

\newpage

# Bibliography