---
title: "Exercise set 8"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: exercises.bib
---

Clear R environment

```{r}
rm(list = ls())
```

# Exercise 8.4

Note: MLE for $\hat\lambda = n/\sum_{i=1}^{n} X_i$, where $X$ denote time
between failure

Refer to the air-conditioning data set **aircondit** provided in the boot
library. The 12 observations are the times in hours between failures of 
air-conditioning equipment

\begin{center}
  \begin{tabular}{ c c c c c c c c c c c c}
   3 & 5 & 7 & 18 & 43 & 85 & 91 & 98 & 100 & 130 & 230 & 487 \\
  \end{tabular}
\end{center}

Assume that the times between failures follow an exponential model with rate
$\lambda$. Obtain the MLE of the hazard rate $\lambda$ and use bootstrap to 
estimate the bias and standard error of the estimate.

in R:

```{r}
library(boot)

# MLE for hazard rate of exponential distributed data
mle <- function(data, i) {
  return(length(data[i])/sum(data[i]))
}

# bootstrapping with 1000 replications
results <- boot(data=aircondit$hours, statistic = mle, R=1000)

# view results
results
plot(results)

# get 95% confidence interval
boot.ci(results, type=c("bca", "norm", "perc"))
```
# Exercise 8.5

\newpage

# Bibliography