---
title: "Exercise set 6"
output: pdf_document
author: Bjørn Christian Weinbach
date: \today
---

Clear R environment

```{r}
rm(list = ls())
```

# Exercise 6.6 in Rizzo

In Exaple 6.7 the control variate approach was illustrated for monte carlo
integration of 
$$
\theta = \int_0^1 e^x dx.
$$

Now consider the antithetic variate approach. Compute $Cov(e^U, e^{1-U})$ and
$Var(e^U + e^{1-U})$, where $U \sim \text{Uniform}(0,1)$. What is the percent
reduction in variance of $\hat{\theta}$ that can be achieved using antithetic 
variates (compared with simple MC)?


```{r}
Nsim <- 10^4
u <- runif(Nsim, 0, 1) 
x1 <- exp(1)^u
x2 <- exp(1)^(1-u)
cov <- cov(x1, x2)
var <- var(x1 + x2)
cov
# finne reduksjon i varians mellom simple MC og antithetic variate MC?
```

## Exercise 6.7 in Rizzo
Refer to exercise 6.6. Use a Monte Carlo simulation to estimate $\theta$ by
the antithetic variate approach and by the simple monte carlo method. Compute
an empirical estimate of the percent reduction in variance using the antithetic 
variate. Compare the result with the theoritical value from Exercise 6.6.

```{r}
nrep <- 1000
intcMCvec <- numeric(nrep)
intaMCvec <- numeric(nrep)
for(i in 1:nrep){
  x <- runif(Nsim)
  intcMCvec[i] <- mean(exp(x)) # crude MC result
  x1 <- runif(Nsim/2)
  x2 <- 1-x1
  x <- c(x1,x2)
  intaMCvec[i] <- mean(exp(x)) # antitetic MC result
}

mean(intcMCvec)
mean(intaMCvec)
1 - (var(intaMCvec) / var(intcMCvec))
```

### Exercise 6.10 in Rizzo
Use MC integration with antithetic variables to estimate
$$
\int_0^1 \frac{e^{-x}}{1 + x^2}dx,
$$
and find the approximate reduction in variance as a percentage of the variance
without variance reduction. 
```{r}
intcMCvec <- numeric(nrep)
intaMCvec <- numeric(nrep)
for(i in 1:nrep){
  x <- runif(Nsim)
  intcMCvec[i] <- mean(exp(-x)/(1 + x^2)) # crude MC result
  x1 <- runif(Nsim/2)
  x2 <- 1-x1
  x <- c(x1,x2)
  intaMCvec[i] <- mean(exp(-x)/(1 + x^2)) # antitetic MC result
}

mean(intcMCvec)
mean(intaMCvec)
1 - (var(intaMCvec) / var(intcMCvec))
```

# Exercise 6.12 in Rizzo
Let $\hat{\theta_f^{IS}}$ be an importans ampling estimator of 
$\theta = \int g(x)dx$, where the importance function f is a density. Prove that 
if $g(x)/f(x)$ is bounded, then the variance of the importance sampling estimator
$\hat{\theta_f^{IS}}$ is finite.

From Rizzo we have

$$
Var(\theta) = E[\hat{\theta}^2] - E[\hat{\theta}]^2 = \int_A \frac{g^2(x)}{f(x)}
- \theta^2
$$
# Problem 1

Calculate the integral
$$
\int_1^\infty \frac{e^{-x}}{1 + x^2} dx
$$
```{r}
Nsim <- 10000
g <- function(x){return(exp(-x) / (1 + x^2) * (x > 1))}
f <- function(x){return(dexp(x))}
x <- rexp(Nsim, 1)
wts <- g(x) / f(x)
intiMC <- mean(wts)
intiMC
```