---
title: "Exercise set 8"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: exercises.bib
---

Clear R environment

```{r}
rm(list = ls())
```

# Exercise 8.4

Note: MLE for $\hat\lambda = n/\sum_{i=1}^{n} X_i$, where $X$ denote time
between failure

Refer to the air-conditioning data set **aircondit** provided in the boot
library. The 12 observations are the times in hours between failures of 
air-conditioning equipment

\begin{center}
  \begin{tabular}{ c c c c c c c c c c c c}
   3 & 5 & 7 & 18 & 43 & 85 & 91 & 98 & 100 & 130 & 230 & 487 \\
  \end{tabular}
\end{center}

Assume that the times between failures follow an exponential model with rate
$\lambda$. Obtain the MLE of the hazard rate $\lambda$ and use bootstrap to 
estimate the bias and standard error of the estimate.

in R:

```{r}
library(boot)

# MLE for hazard rate of exponential distributed data
mle <- function(data, i) {
  return(length(data[i])/sum(data[i]))
}

# bootstrapping with 1000 replications
results <- boot(data=aircondit$hours, statistic = mle, R=1000)

# view results
results
plot(results)

# get 95% confidence interval
boot.ci(results, type=c("bca", "norm", "perc"))
```
# Exercise 8.5
Refer to exercise 8.4. Compute $95\%$ confidence interval for the mean time 
between failures by the standard normal, basic, percentile and BCa methods.

```{r}
# MLE for hazard rate of exponential distributed data
meantimeest <- function(data, i) {
  rate <- length(data[i])/sum(data[i])
  return(1/rate)
}

# bootstrapping with 1000 replications
results <- boot(data=aircondit$hours, statistic = meantimeest, R=1000)

# view results
results
plot(results)

# get 95% confidence interval
boot.ci(results, type=c("norm", "basic", "perc", "bca"))
```

# Exercise 11.3

Use metropolis-hastings sampler to generate random variables from a standard
Cauchyy distribution. Discard the first 1000 of the chain, and compare the 
deciles of the generated observations with the deciles of the standard Cauchy
distribution. Recall that a $\text{Cauchy}(\theta, \eta)$ has density

\begin{equation}
f(x) = \frac{1}{\theta \pi(1 + [(x - \eta)/\theta]^2)}, 
  \quad -\infty < x < \infty, \quad \theta > 0.
\end{equation}

In R:

```{r}
# general 1d Gaussian proposal random walk MH
oneD.RWMH <- function(lprob,
                     sigma=1.0,
                     theta1=0.0,
                     n.iter=10000){
  # space for output
  output <- numeric(n.iter)
  # first iterate given
  output[1] <- theta1
  # Calculate probability of theta 1
  lp.old <- lprob(theta1)
  # main iteration loop
  for(t in 2:n.iter){
    # proposal
    thetaStar <- output[t-1] + rnorm(1,sd=sigma)
    # accept probability, for numerical stability we compute
    lp.star <- lprob(thetaStar)
    # the log-accept prob, and then take exp
    alpha <- exp(min(0.0,lp.star-lp.old))
    # accept/reject step
    if(runif(1)<alpha && is.finite(alpha)){
      output[t] <- thetaStar
    } else {
      output[t] <- output[t-1]
    }
  }
print(paste0("RWMH done, accept prob : ",mean(abs(diff(output))>1.0e-14)))
  return(output[1000:length(output)])
} 


std_cauchy <- function(x){return(dcauchy(x, 0, 1, log=TRUE))}

# try algorithm (change initial condition and sigma for illustration)
out <- oneD.RWMH(std_cauchy, theta1 = 0, sigma=2)
par(mfrow=c(2,1))
plot(1:length(out),out,pch=20,cex=0.1,xlab="MCMC iteration #")
hist(out)
```

This implementation uses standard normal proposals. We can see from the plot that the random walk stops suddenly on values
far from the center of teh distribution and skews the samples. Let's perform some diagnostics:

```{r}
#
# convergence tests 
#
library(coda)
#
# Geweke (single chain test)
#
# make an mcmc object used by the coda package
chain.mcmc <- mcmc(data=out)
varnames(chain.mcmc) <- "theta"
# returns z-score (to be compared two-sided with standard normal)
geweke.diag(chain.mcmc)
# check if we should discard more burn-in
geweke.plot(chain.mcmc) 

```
\newpage

# Bibliography