---
title: "Exercise set 5"
output: pdf_document
author: Bjørn Christian Weinbach
date: \today
---

### Clear R environment

```{r}
rm(list = ls())
```

# Exercise 4.3

A nonhomogeneous Poisson process has the mean value function

$$
m(t) = t^2 + 2t, \qquad t \geq 0.
$$

Determine the intensity function $\lambda(t)$ of the process, and write
a program to simulate the process on the interval $[4,5]$. Compute the probability distribution of  $N(5) - N(4)$, and compare it to the empirical estimate obtained by replicating the simulation.

```{r}
# Function for simulating arrival times for a NHPP between a and b using thinning
simtNHPP <- function(a,b,lambdamax,lambdafunc){
  if(max(lambdafunc(seq(a,b,length.out = 100)))>lambdamax)
    stop("lambdamax is smaller than max of the lambdafunction")
  expectednumber <- (b-a)*lambdamax  
  Nsim <- 3*expectednumber
  timesbetween <- rexp(Nsim,lambdamax) 
  timesto <- a+cumsum(timesbetween)
  timesto <- timesto[timesto<b]
  Nevents <- length(timesto)
  U <- runif(Nevents)
  timesto <- timesto[U<lambdafunc(timesto)/lambdamax]  
  timesto
}

# Specify the intensity function for the traffic example
ex43 <- function(t)
  2*t + 2

tvec <- seq(4, 5,by=0.01)
plot(tvec, ex43(tvec),type="l")
```

```{r}
# Generate data with the traffic intensity and plot them
NHPPtimes <- simtNHPP(a=4,b=5,lambdamax=12,lambdafunc=ex43)
plot(NHPPtimes,1:length(NHPPtimes),type="s",xlab = "time", 
     ylab = "Event number",lwd=1.5)
points(NHPPtimes,rep(0,length(NHPPtimes)),pch=21,bg="red")
# Rerun the lines above several times
```


# Exercise 6.1

Compute a Nonte Carlo estimate of 

$$
\int_0^{\frac{\pi}{3}} sin(t) dt
$$

```{r}
mcint <- function(Nsim, a, b, func) {
  return((b - a)*mean(func(runif(Nsim, a, b))))
}

intfunc <- function(x) {
  return(sin(x))
}

print(paste("Monte Carlo Integration:", mcint(10000, 0, pi/3, intfunc)))
print(paste("Exact integral:", -cos(pi/3) + cos(0)))
```

# Exercise 6.2

Refer to example 6.3. Compute a Monte Carlo estimate of the standard normal cdf,
by generating from the uniform(0, x) distribution. Compare estimate with normal
cdf function pnorm.

```{r}
intfunc <- function(t) {
  return(exp(-t^2/2) * sqrt(2/pi))
}

theta <- mcint(10000, 0, 2, intfunc)

print(paste("Monte Carlo Integration:", theta))
print(paste("Exact integral:", pnorm(2, 0, 1, lower.tail = TRUE, log.p = FALSE)))
```

## Calculate confidence interval

```{r}
theta
sdtheta <- sd(exp(-runif(10000, 0, 2)^2 / 2) * sqrt(2 / pi))
theta - qnorm(0.975, 0, 1)*sdtheta
theta + qnorm(0.975, 0, 1)*sdtheta
```

# Exercise 6.3

Compute a Monte Carlo estimate $\hat{\theta}$ of

$$
\int_{0}^{0.5} e^{-x} dx
$$

by sampling from a Uniform(0, 0.5) and estimate the variance of of $\hat{\theta}$

```{r}
Nsim <- 10000
thetahat <- (0.5 - 0)*mean(exp(-runif(Nsim, 0, 0.5)))
thetastar <- pexp(0.5, 1) - pexp(0, 1)

print(paste("Monte Carlo Integration:", mcint(10000, 0, 0.5, intfunc)))
print(paste("Exponential distribution:", thetastar))
```

