---
title: "Exercise set 4"
output: pdf_document
author: Bjørn Christian Weinbach
date: \today
---

# Exercise 3.1
I use inverse transform sampling to sample numbers from an two-parameter
exponential distribution. We have from example 2.6 in Rizzo that the cdf is:

\begin{equation}
F_X(x) = 1 - e^{-\lambda(x - \eta)} , \qquad x \geq \eta
\end{equation}

Which gives us

\begin{equation}
F^{-1}_X(U) = \frac{\eta\lambda - ln(1 - U)}{\lambda}, 
\qquad U \in \text{Uniform}(0, 1)
\end{equation}

That has the same distribution as X.

Below is the R Code for sampling numbers from the distribution:

### Inverse cumulative function.

```{r}
# itm (Inverse Transform Method) function for equation one in assignment 1
itm_twoexp <- function(n, lambda, eta) {
  u <- runif(n)
  return (((lambda*eta) - (log(1 - u)) ) / lambda)
}
```

### Sample $Nsim$ numbers and plot in histogram.
```{r}
Nsim <- 10000000
lambda <- 4
eta <- 2
itm_p <- itm_twoexp(Nsim, lambda, eta)

hist(itm_p, prob = TRUE, xlab = "X", ylab = "P(X)",
     main="Histrogram of inverse transform method")
lines(density(itm_p))
```


# Exercise 3.11
We have:
\begin{itemize}
  \item $N = 1000$
  \item $p_1$
  \item $p_2 = 1 - p_1$ 
  \item $N_1(0, 1)$
  \item $N_2(3, 1)$
\end{itemize}

Where $p_1$ and $p_2$ are probabilities of mixing the two distributions.
This has been implemented in R like so:
```{r}
mixture <- function(n, u1, s1, u2, s2, p) {
  n <- n
  x1 <- rnorm(n, u1, s1)
  x2 <- rnorm(n, u2, s2)
  u <- runif(n)
  k <- as.integer(u > p)
  x <- k * x1 + (1-k) * x2
  hist(x, prob = TRUE, xlab = "X", ylab = "P(X)",
       main=paste("N(0,1), N(3,1), p=", round(p, 2)))
  lines(density(x))
}

par(mfrow=c(2,3))
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 0/5)
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 1/5)
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 2/5)
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 3/5)
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 4/5)
mixture(n = 1000, u1 = 0, s1 = 1, u2 = 3, s2 = 1, p = 5/5)
par(mfrow=c(1,1))
```

We have that $p_1 \in [0, 1]$ as it is a probability. When we plot the 
histograms for different values of $p_1$ ranging from $0$ to $1$ we observe 
that when $p_1 = 0$ we get only values from  
$N(0, 1)$ and vice versa. In either case, the distribution is no longer bimodal. 

With this observation i present the following conjecture:

\newtheorem{conjecture}{Conjecture}

\begin{conjecture}
Given a normal mixture $X$, denoted $F_X(x) = (1 - p)F_{X_1}(x) + pF_{X_2}(x)$.
Approaches a normal distribition in two cases

\begin{enumerate}
  \item \begin{equation}
          \lim_{p \to 0} F_X(x) \rightarrow X_1 \sim N(\mu_1, \sigma_1)
        \end{equation}
  \item \begin{equation}
        \lim_{p \to 1} F_X(x) \rightarrow X_2 \sim N(\mu_2, \sigma_2)
      \end{equation}
\end{enumerate}
\end{conjecture}

# Exercise 3.14
In this exercise we are tasked with generating 200 random observations from
a 3-dimensional multivariate normal distribution with mean vector
$\mu = (0, 1, 2)$ and covariance matrix
\begin{equation*}
\Sigma = \begin{bmatrix}
          1.0 & -0.5 & 0.5\\
          -0.5 & 1.0 & -0.5 \\
          0.5 & -0.5 & 1.0
          \end{bmatrix}
\end{equation*}

## Choleski factorization  

```{r}
rmvn.Cholenski <-  function(n, mu, sigma) {
  # Generate n random vectors from MVN(mu, Sigma)
  # dimensions are inferred from mu and Sigma
  d <- length(mu)
  Q <- chol(sigma) # Cholenski factorization of sigma
  Z <- matrix(rnorm(n*d), nrow = n, ncol = d)
  X <- Z %*% Q + matrix(mu, n, d, byrow = TRUE)
}

n <- 200
mu <- c(0, 1, 2)
sigma <- matrix(
            c(1.0, -0.5, 0.5, -0.5, 1.0, -0.5, 0.5, -0.5, 1.0), 
            nrow = 3, 
            ncol = 3
        )

X <- rmvn.Cholenski(200, mu, sigma)
pairs(X)
```


# Exercise 1

## Simulate arrival times with the fact that no of arrivals is $\lambda t$

```{r}
# implementation based on the relation to the uniform distribution
plotHPPu <- function(lambda, t){
  Nevents <- rpois(1, lambda * t)
  timesto <- sort(runif(Nevents, min=0, max=t))
  
  plot(timesto, 1:Nevents, type="s", xlab = "arrival time", 
       ylab = "Event number", lwd=1.5, ylim=c(0, Nevents))
  
  points(timesto, rep(0,Nevents), pch=21, bg="red")
}

plotHPPu(lambda = 2, t = 10)
```

## Calculate $P(N(10) > 25)$

The probability of $P(N(10) > 25)$ and the poisson process has the property
that the number of events is poisson distributed. Which means that the probability
is the same as the probability of a poisson distributed random variable $X$
with parameter $\lambda t = 2 \times 10 = 20$ is larger than $25$ i.e
$$P(X > 25) = 1 - P(X \leq 25) = 1 - e^{-\lambda t}\sum_{i=0}^{25} \frac{(\lambda t)^i}{i!}$$

## Calculate above equation by simulation

```{r}
rHPPu <- function(Nsim, lambda, t){
  Nevents <- rpois(Nsim, lambda * t)
  return(mean(Nevents > 25))
}

lambda <- 2
t <- 10
rHPPu(1000000, lambda, t)
```

## Analytical calculation

```{r}
ppois(q = 25, lambda = lambda*t, lower.tail = FALSE, log.p = FALSE)
```