---
title: "Exercise set 7"
author: "Bjørn Christian Weinbach"
date: \today
header-includes:
   - \usepackage{bbm}
   - \usepackage[UKenglish]{babel}
   - \usepackage[T1]{fontenc}
   - \usepackage[nodayofweek,level]{datetime}
urlcolor: blue
output: pdf_document
bibliography: exercises.bib
---

Clear R environment

```{r}
rm(list = ls())
```

# Problem 0
$N = 10$ measurments of the measurements of the hardness of a new material 
(alloy) gave the following data:

\begin{center}
  \begin{tabular}{ c c c c c }
   168 & 185 & 164 & 182 & 169 \\
   181 & 172 & 185 & 172 & 180 \\  
  \end{tabular}
\end{center}

A typical approach is to assume that the hardness measurements follow a 
$N(\mu; \sigma^2)$ distribution.

## Calculate the empirical mean $\hat{\mu}$ and standard deviation
$\hat{\sigma}$.

```{r}
data <- c(168, 185, 164, 182, 169, 181, 172, 186, 172, 180)

print(paste("Empirical mean: ", mean(data)))
print(paste("Empirical sd: ", sd(data)))
```

## Calculate the standard deviation of $\hat{\mu}$

According to the wikipedia article on standard error accessed 
\formatdate{22}{10}{2020} [@wiki:SampleMean] For each random variable, the sample 
mean is a good estimator of the population 
mean, where a "good" estimator is defined as being efficient and unbiased. 
Of course the estimator will likely not be the true value of the population 
mean since different samples drawn from the same distribution will give 
different sample means and hence different estimates of the true mean. 
Thus the sample mean is a random variable, not a constant, 
and consequently has its own distribution.

We can therefore calculate the standard deviation of $\hat{\mu}$, also called
the standard error of the mean as follows:

\begin{equation}
  \sigma_{\bar{X}} = \frac{\sigma}{N}
\end{equation}[@wiki:StandardError]

Since the population mean is seldom known. We can estimate this like so:

\begin{equation}
  \sigma_{\bar{X}} \approx \frac{s}{N}
\end{equation}

Where $s$ is the sample standard deviation.

In R:

```{r}
print(paste("Standard error: ", sd(data)/sqrt(length(data))))
```
## Calculate $95\%$ confidence intervals for $\mu$ and $\sigma$

Since we have a unknown $\mu$ and $\sigma$ for the population and the sample
size is small, we will use a T values for our confidence interval for the population mean.

The confidence interval is
\begin{equation}
[\bar{x} + t_{n-1, \alpha/2} \cdot \frac{s}{\sqrt{n}}, 
 \bar{x} + t_{n-1, 1 - \alpha/2} \cdot \frac{s}{\sqrt{n}}]
\end{equation}

in R:

```{r}
# Calculate sample mean, sd and no of observations
xbar <- mean(data)
s <- sd(data)
n <- length(data)

# Alpha
a <- 0.05

# Calculate confidence interval. qt is T-distribution
left <- xbar + qt(a/2, n-1) * s/sqrt(n)
right <- xbar + qt(1 - (a/2), n-1) * s/sqrt(n)
ci <- c(left, right)

# Display confidence interval
ci
```
To calculate the confidence interval for $\sigma$ we use the formula in the 
text of the exercise set. 

```{r}
left <- sqrt((n-1)*s^2 / qchisq(1 - a/2, df=n-1))
right <- sqrt((n-1)*s^2 / qchisq(a/2, df=n-1))
ci <- c(left, right)
ci
```
## Bootstrapping $\hat{\mu}$

Now we resort to bootstrapping to estimate the calculations we have above from our data.
We first do this for the statistic $\hat{\mu}$


```{r}
library(boot)

# Function for statistic - sample mean
meanestfunc <- function(data,i)
  mean(data[i])

boot.obj <- boot(data=data, statistic = meanestfunc, R=5000)
boot.obj
boot.ci(boot.obj,type=c("norm","basic","perc","bca"))
```
We see that our calculations for the standard error is close to the estimated 
standard error by bootstrapping. The confidence interval is also close to our interval. 
Though the calculated one used t-values and is therefore slightly wider.

## Bootstrapping $\hat{\sigma}$

```{r}
# Function for statistic - sample mean
stdfunc <- function(data,i)
  sd(data[i])

boot.obj <- boot(data=data, statistic = stdfunc, R=5000)
boot.obj
boot.ci(boot.obj,type=c("norm","basic","perc","bca"))
```

\newpage

# Bibliography